---
title: "Benchmarks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Warning

This vignette has not been updated yet for luajr 0.2.2.9000 (in development).

In the same way that using R packages 
[`cpp11`](https://CRAN.R-project.org/package=cpp11) or 
[`Rcpp`](https://CRAN.R-project.org/package=Rcpp) allows you to 
bridge your R code with C++ code that runs faster than R, you can use `luajr` 
to bridge your R code with Lua code that runs faster than R.

In general, C/C++ code runs about 5-1,000 times faster than the equivalent R 
code.

In my experience, `luajr` code often presents a similar speedup, about the 
same as C/C++ code, or in the worst case, maybe half as fast. Sometimes 
`luajr` code can be faster than C/C++, though usually it isn't quite as good. 

Why use `luajr` then? `Rcpp` and `cpp11` require a C++ toolchain (e.g. 
gcc, clang, etc.) and requires long compilation times, whereas `luajr` 
doesn't. This means that `luajr` is usable when a C++ compiler isn't 
available, or when compilation times are prohibitive or an annoyance.

In the following section, we look at two aspects of benchmarking. In the first
example, we compare different ways of passing vectors to Lua functions relative
to R. In the second example, we compare more fundamentally the difference in
running a whole algorithm in Lua versus R versus C++ with `Rcpp`. 

### Parameter passing example: sum of squares

For reasonably long vectors, passing by reference (`'r'`) is faster than passing 
by value (`'v'`), which is faster than passing by simplify (`'s'`, `'a'`, or 
`'1'`&ndash;`'9'`). But for relatively short vectors, passing by simplify can 
avoid some overhead, and for very simple operations on very short vectors, it 
might not be worth using Lua at all.

To illustrate the above points, here is some code that takes a numeric vector 
and calculates the sum of squares, comparing a pure R function with three 
alternatives written in Lua, respectively passing the vector by reference, by 
value, and by simplify.

```{r}
library(luajr)

v1 = rnorm(1e1)
v4 = rnorm(1e4)
v7 = rnorm(1e7)

lua("sum2 = function(x) local s = 0; for i=1,#x do s = s + x[i]*x[i] end; return s end")
sum2 = function(x) sum(x*x)
sum2_r = lua_func("sum2", "&.")
sum2_v = lua_func("sum2", ".")
sum2_s = lua_func("sum2", "$.")

# Comparing the results of each function:
sum2(v1)    # Pure R version
sum2_r(v1)  # luajr pass-by-reference
sum2_v(v1)  # luajr pass-by-value
sum2_s(v1)  # luajr pass-by-simplify
```

The time taken to sum `v1`, `v4` and `v7`, depending upon the function kind, on 
a 2019-era MacBook Pro is summarised in this table:

|Call            |Number of summands | Method                    | Median runtime |
|:---------------|:-----------------:|:-------------------------:|---------------:|
|`sum2(v1)`      | 10                | Pure R                    |       **1 µs** |
|`sum2_r(v1)`    | 10                | `luajr` pass by reference |         6 µs   |
|`sum2_v(v1)`    | 10                | `luajr` pass by value     |         7 µs   |
|`sum2_s(v1)`    | 10                | `luajr` pass by simplify  |         4 µs   |
|                |                   |                           |                |
|`sum2(v4)`      | 10,000            | Pure R                    |        63 µs   |
|`sum2_r(v4)`    | 10,000            | `luajr` pass by reference |      **18 µs** |
|`sum2_v(v4)`    | 10,000            | `luajr` pass by value     |        93 µs   |
|`sum2_s(v4)`    | 10,000            | `luajr` pass by simplify  |        78 µs   |
|                |                   |                           |                |
|`sum2(v7)`      | 10,000,000        | Pure R                    |    49,000 µs   |
|`sum2_r(v7)`    | 10,000,000        | `luajr` pass by reference |  **13,000 µs** |
|`sum2_v(v7)`    | 10,000,000        | `luajr` pass by value     |    96,000 µs   |
|`sum2_s(v7)`    | 10,000,000        | `luajr` pass by simplify  |   124,000 µs   |

When the vector is 10 elements long, the R version wins handily.

When the vector is 10,000 elements long, the pass-by-reference Lua version is
fastest, but the other methods are all comparable.

When the vector is 10,000,000 elements long, the story is similar. 

This is an example where `luajr` doesn't add that much speed&mdash;the 
function is relatively short, and there is a certain amount of overhead in 
invoking it, while the R function doesn't have the overhead of transferring 
control between languages.

### Processing time example: logistic map

Consider the following:

```{r}
logistic_map_R = function(x0, burn, iter, A)
{
    result_x = numeric(length(A) * iter)
    
    j = 1
    for (a in A) {
        x = x0
        for (i in 1:burn) { 
            x = a * x * (1 - x)
        }
        for (i in 1:iter) { 
            result_x[j] = x
            x = a * x * (1 - x)
            j = j + 1
        }
    }
    
    return (list2DF(list(a = rep(A, each = iter), x = result_x)))
}

logistic_map_L = lua_func(
"function(x0, burn, iter, A)
    local dflen = #A * iter
    local aa = luajr.numeric(dflen, 0)
    local xx = luajr.numeric(dflen, 0)
    
    local j = 1
    for k,a in pairs(A) do
        local x = x0
        for i = 1, burn do
            x = a * x * (1 - x)
        end
        for i = 1, iter do
            aa[j] = a
            xx[j] = x
            x = a * x * (1 - x)
            j = j + 1
        end
    end
    
    local result = luajr.dataframe()
    result:set('a', aa)
    result:set('x', xx)
    
    return result
end", "$., $., $., .")

# To be compiled using Rcpp::cppFunction()
logistic_map_C =
'DataFrame logistic_map(double x0, unsigned int burn, unsigned int iter, NumericVector A)
{
    unsigned int dflen = A.length() * iter;
    NumericVector da(dflen, 0);
    NumericVector dx(dflen, 0);
    
    unsigned int j = 0;
    for (auto a : A)
    {
        double x = x0;
        for (unsigned int i = 0; i < burn; ++i)
            x = a * x * (1 - x);
        for (unsigned int i = 0; i < iter; ++i, ++j)
        {
            dx[j] = x;
            da[j] = a;
            x = a * x * (1 - x);
        }
    }

    return DataFrame::create(Named("a") = da, Named("x") = dx);
}'
```

Here we are comparing three different versions (R, Lua, C++) of running a 
parameter sweep of the logistic map, a chaotic dynamical system popularized by 
Bob May in a [1976 Nature article](https://abel.math.harvard.edu/archive/118r_spring_05/docs/may.pdf).
The output looks like this:

```{r}
#| fig.alt: >
#|   Plot of fixed points of the logistic map chaotic dynamical system.
#|   The input x-axis value a ranges from 2.0 to 3.85. Between 2.0 and 
#|   3.0 there is a single fixed point; above 3.0 the fixed points exhibit
#|   bifurcations and chaotic behaviour.
logistic_map = logistic_map_L(0.5, 100, 100, 200:385/100)
plot(logistic_map$a, logistic_map$x, pch = ".")
```

In the code above, we don't actually compile the C++ code using `Rcpp`, 
we only declare the source code as an illustration. I don't actually compile 
the code in this article because I didn't want to have `luajr` depend on 
`Rcpp` only for the sake of this vignette, but you would do so using
`Rcpp::cppFunction(logistic_map_C)`.

The times taken by each function on a 2025-era MacBook Pro are as follows:

|Call                                          | Method       | Median runtime |
|----------------------------------------------|--------------|---------------:|
|`logistic_map_R(0.5, 100, 100, 200:385/100))` | R function   |       850 µs   |
|`logistic_map_L(0.5, 100, 100, 200:385/100))` | Lua function |      **80 µs** |
|`logistic_map_C(0.5, 100, 100, 200:385/100))` | C++ function |       100 µs   |

The version written in Lua is around 10 times faster than the version in R, 
and even somewhat outperforms `Rcpp`. Note that the relative speed of Lua 
versus C++ depends on the number of iterations, and whether you use `Rcpp` or
`cpp11`, which each seem to have marginal advantages in different cases. 
Nonetheless, the Lua version seems to execute about as quickly as the C++ 
version, within plus or minus 20%.

The speedup of the Lua function relative to R was much more notable in an 
earlier test where the R version first created the data frame and then 
performed the iteration, i.e. with the line `result$x[j] = x` instead of 
`result_x[j] = x`. The median runtime for that R version was two orders of 
magnitude slower than the Lua version; the extra overhead associated with 
`data.frame` methods was pointed out by Tim Taylor.

